---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

The following dependencies exist for this report.
```{r dependencies, echo=TRUE, warning=FALSE, message=FALSE }
library(data.table)
library(printr)
library(lubridate)
library(lattice)
```

## Loading and preprocessing the data

Read the data in from the files and save it into a data.table
```{r load-data }
activity <- read.csv("activity.csv", na.strings = "NA")
activity.dt <- data.table(activity)
```

## What is the mean total number of steps taken per day

1. Calculate the total number of steps per day
```{r total-steps}
# remove the na lines
activity.dt.noNA <- na.omit(activity.dt)
# sum steps by date
activity.dt.steps.by.day <- activity.dt.noNA[,list(steps.total=sum(steps)),by=date]
```
Total steps by date:
`r activity.dt.steps.by.day`

2. Histogram of step by day

```{r, echo=FALSE}
  hist(activity.dt.steps.by.day$steps.total, xlab="Steps per Day",
   main="Histogram of Steps per Day")
```

3. Calculate the mean and the median of the total number of steps per day
```{r calc-step-stats}
step.stats<-activity.dt.steps.by.day[,list(total.mean=mean(steps.total),total.median=median(steps.total))]
```
Step stats - mean: `r format(round(step.stats$total.mean, 3), nsmall=2)`,   median: `r step.stats$total.median`

## What is average daily activity pattern?

We want to see how much exercise people do on average during each time interval.

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r calc-steps-by-interval}
  # formatter to convert 815 to "08:15" and 2255 to "22:55"
intToHHMM <- function (i) {  sprintf("%02d:%02d", i %/% 100, i %% 100)}
# calculate the mean steps for each time interval
activity.dt.steps.by.interval.mean <- activity.dt.noNA[,list(steps.mean=mean(steps)),by=interval]
# add the column "interval.HHMM" which displays the interval as a string in the format HH:MM
activity.plot <- activity.dt.steps.by.interval.mean[, c("interval.HHMM") :=intToHHMM(interval), key="interval"]
# generate the plot
plot(strptime(activity.plot[, interval.HHMM], format="%H:%M"), activity.plot[, steps.mean],type="l", xlab="Time of Day (HH:MM)", ylab="Number of steps (Mean)", main="Average Activity by Time")
```


2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r calc-steps-by-interval.maxstepinterval}
max.steps.row<-activity.plot[,.SD[which.max(steps.mean)]]
```
Interval `r max.steps.row$interval` (`r max.steps.row$interval.HHMM`) has the average maximum number of steps: `r max.steps.row$steps.mean `


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r Missing values}
numMissing <- nrow(activity.dt) - nrow(activity.dt.noNA)
```
There were `r numMissing` rows with missing values

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

* Using the median value for the interval to fill in the missing data

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r Fillin missing values}
# for each interval with a missing reading use the mean for that interval as the value
#   Create a copy of dt with NAs (this is ineffiecient, but I do not want to lose things while experimenting)
activity.dt.filled <- activity.dt
#   Get medians for each interval
activity.dt.steps.by.interval.median <- activity.dt.noNA[,list(steps.median=median(steps)),by=interval]
#   replace steps with NA with the median.  This would be more efficent with get() calls.
activity.dt.filled$steps <- ifelse(is.na(activity.dt.filled$steps), activity.dt.steps.by.interval.median$steps.median[match(activity.dt.filled$interval,activity.dt.steps.by.interval.median$interval)], activity.dt.filled$steps)

```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r }
# sum filled steps by date
activity.dt.filled.steps.by.day <- activity.dt.filled[,list(steps.filled.total=sum(steps)),by=date]
# create comparison of daily steps without NAs and with NAs imputed
activity.dt.compare <- merge (activity.dt.steps.by.day, activity.dt.filled.steps.by.day, by='date',all=TRUE)
```
* Compare the total steps by date without NAs and with NAs filled:
`r activity.dt.compare`

**  As the table shows the effect is to fill in the and thus add low # of step rows to the data set.

* Histogram of step by day

```{r, echo=FALSE}
# to do replace with a melt and ggplot2
  hist.noNA <- hist(activity.dt.steps.by.day$steps.total, plot=FALSE)
  hist.data.filled = hist(activity.dt.filled.steps.by.day$steps.filled.total, plot=FALSE)
  # deterine ranges
  #xlim <- range(activity.dt.steps.by.day$date, activity.dt.filled.steps.by.day$date)
  ylim <- range(0, activity.dt.steps.by.day$steps.total, activity.dt.filled.steps.by.day$steps.filled.total )
  # plot 1st graph2
  plot(hist.noNA, 
     col = rgb(1,0,0,0.4),xlab = 'Steps per day',
     main = 'Histogram of steps per day')
  # plot of 2nd graph
  opar <- par(new = FALSE)
  plot(hist.data.filled, 
     xaxt = 'n', yaxt = 'n', ## don't add axes
     col = rgb(0,0,1,0.4), add = TRUE)
  ## add a legend in the corner
  legend('topleft',c('NA Removed','NA filled'),
       fill = rgb(1:0,0,0:1,0.4), bty = 'n',
       border = NA)
  par(opar)
```

  **  The histogram shows how the low step values have been shifted.


* Calculate the mean and the median of the total number of steps per day
```{r calc-step-stats-filled}
step.filled.stats<-activity.dt.filled.steps.by.day[,list(total.mean=mean(steps.filled.total),total.median=median(steps.filled.total))]
```
Step stats (with NAs) - mean: `r format(round(step.stats$total.mean, 3), nsmall=2)`,   median: `r step.stats$total.median`
Step stats (without NAs) - mean: `r format(round(step.filled.stats$total.mean, 3), nsmall=2)`,   median: `r step.filled.stats$total.median`

  **  The step stats (mean and median) show a downward shift in resulting from weighting the data at the low end of the step count.


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r add and populate Day of Week column }
#   create new data.table, activity.dt.filled.DoW
activity.dt.filled.DoW <- activity.dt.filled
#   Add a column for the type of day ("Weekend", "Weekday")
activity.dt.filled.DoW$typeOfDay <- ifelse(wday(as.Date(activity.dt.filled$date)) ==1 | wday(as.Date(activity.dt.filled$date)) ==7  , "Weekend", "Weekday")
#   Make the column a factor
activity.dt.filled.DoW$typeOfDay <- as.factor(activity.dt.filled.DoW$typeOfDay)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r }
# calculate the mean steps for each time interval
activity.plot.DoW <- activity.dt.filled.DoW[,list(steps.mean=mean(steps)),by=c('interval', 'typeOfDay')]
#activity.dt.filled.DoW.steps.by.interval.mean
# add the column "interval.HHMM" which displays the interval as a string in the format HH:MM
activity.plot.DoW <- activity.plot.DoW[, c("interval.HHMM") :=intToHHMM(interval), key="interval"]
# plot by type of day to compare weeekday to weekend activities
#    get the max y values so we can use it to ensure the y ranges are the same.
steps.mean.max <- max(activity.plot.DoW$steps.mean)
#    plot one plot on top of the other
par(mfrow=c(2,1))
#    plot the Weekday
activity.plot.DoW.Weekday <- activity.plot.DoW[typeOfDay == "Weekday",]
plot(strptime(activity.plot.DoW.Weekday[, interval.HHMM], format="%H:%M"), activity.plot.DoW.Weekday[, steps.mean],type="l", xlab="Time of Day (HH:MM)", ylab="Number of steps (Mean)", ylim=c(0, steps.mean.max), main="Weekday Average Activity by Time")
#    plot the Weekend
activity.plot.DoW.Weekend <- activity.plot.DoW[typeOfDay == "Weekend",]
plot(strptime(activity.plot.DoW.Weekend[, interval.HHMM], format="%H:%M"), activity.plot.DoW.Weekend[, steps.mean],type="l", xlab="Time of Day (HH:MM)", ylab="Number of steps (Mean)", ylim=c(0, steps.mean.max), main="Weekend Average Activity by Time")

```

** The plots show that there is higher peek activity during weekdays and activity starts to build earlier.  On the weekend the 
activity starts later, but is higher throughout the middle of the day and continues later into the evening.   This is 
likely reflects that during weekdays people go to work in the morning and settle in for the day.  On weekends they get up later but
are active throughout the day and stay active later. 
